dispRity manual

### 2017-08-12 gitbook/index.Rmd

---
title: "dispRity manual"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 2
bibliography: References.bib
bst: sysbio.bst
vignette: >
  %\VignetteIndexEntry{dispRity manual}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

@@

# Introduction

@@

# Getting started

## What sort of data does dispRity work with?

Disparity can be estimated from pretty much any matrix.
Classically, however, it is measured from ordinated matrices.
These matrices can be from any type of ordination (PCO, PCA, PCoA, MDS, etc.) as long as they have your element names as rows (taxa, experiments, countries etc.) and your ordination axes as columns (the dimensions of your dataset).

### Ordination matrices from `Claddis`

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if Claddis is installed BUILD ONLY
if(!require(Claddis)) install.packages("Claddis")
```

`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:

```{r}
require(Claddis)

## Ordinating the example data from Claddis
Claddis.ordination(Michaux1989) 
```

Note that several options are available, namely which type of distance should be computed.
See more info in the function manual (`?Claddis.ordination`).
Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.

### Ordination matrices from `geomorph`

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if geomorph is installed BUILD ONLY
if(!require(geomorph)) install.packages("geomorph")
```

You can also easily use data from `geomorph` using the `geomorph.ordination` function.
This function simply takes Procrustes aligned data and performs an ordination:

```{r}
require(geomorph)

## Loading the plethodon dataset
data(plethodon)

## Performing a Procrustes transform on the landmarks
procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)

## Ordinating this data
geomorph.ordination(procrustes)[1:5,1:5]
```

Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
Additionally you can give the function a `geomorph.data.frame` object.
If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsamples)!

```{r}
## Using a geomorph.data.frame
geomorph_df <- geomorph.data.frame(procrustes,
     species = plethodon$species, site = plethodon$site)

## Ordinating this data and making a dispRity object
geomorph.ordination(geomorph_df)
```

More about these `dispRity` objects below!

### Other kinds of ordination matrices

If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
Here is how to do it for the following types of matrices:

 * Multivariate matrices (principal components analysis; PCA)

```{r}
## A multivariate matrix
head(USArrests)

## Ordinating the matrix using `prcomp` 
ordination <- prcomp(USArrests)

## Selecting the ordinated matrix
ordinated_matrix <- ordination$x
head(ordinated_matrix)
```
  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.

 * Distance matrices (classical multidimensional scaling; MDS)

```{r}
## A matrix of distances between cities
str(eurodist)

## Ordinating the matrix using cmdscale() with k = 5 dimensions 
ordinated_matrix <- cmdscale(eurodist, k = 5)
head(ordinated_matrix)
```

  This results in a ordinated matrix with European cities as elements and five dimensions.

Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.

## Performing a simple dispRity analysis

Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).

### Example data

To illustrate these functions, we will use data from @beckancient2014.
This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).

```{r, fig.width=7, fig.height=7}
## Loading the ordinated matrices
data(BeckLee_mat50)
data(BeckLee_mat99)

## The first five taxa and dimensions of the 50 taxa matrix
head(BeckLee_mat50[, 1:5])

## The first five taxa and dimensions of the 99 taxa + ancestors matrix
BeckLee_mat99[c(1, 2, 98, 99), 1:5]

## Loading a list of first and last occurrence dates for the fossils
data(BeckLee_ages)
head(BeckLee_ages)

## Loading and plotting the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) 
axisPhylo(root = 140)
nodelabels(cex = 0.5)
```

Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).

### Disparity through time

The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
Of course all the defaults can be changed if required, more on this later.

For a disparity through time analysis, you will need:

  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time (i.e. an age for the root), you can simply do `my_tree$root.time <- my_age`.
  * The required number of time subsamples (here `time = 3`)
  * Your favourite disparity metric (here the sum of variances)

Using the @beckancient2014 data described [above](#example-data):

```{r}
## Measuring disparity through time
disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
                                        time = 3, metric = c(sum, variances))
```

This generates a `dispRity` object (see [here](#manipulating-dispRity-objects) for technical details).
When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:

```{r}
## Print the disparity_data object
disparity_data
```

We asked for three subsamples (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.

We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`):

```{r, fig.width=7, fig.height=7}
## Summarising disparity through time
summary(disparity_data)

## Plotting the results
plot(disparity_data, type = "continuous")

## Testing for an difference among the time bins
disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
summary(disp_lm)

## Note that ANOVA assumptions were not checked here!
```

Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
You can also directly explore the specific function help files within R and navigate to related functions.

### Disparity among groups

The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
For example, you could ask if there is a difference in disparity between two groups?

To perform such an analysis, you will need:

 * An matrix with rows as elements and columns as dimensions (always!)
 * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix. For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4. Note that elements can be present in multiple groups at once.
 * Your favourite disparity metric (here the sum of variances)

Using the @beckancient2014 data described [above](#example-data):

```{r}
## Creating the two groups (crown versus stem) as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44))

## Measuring disparity for each group
disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
                                     metric = c(sum, variances))
```

We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).

```{r, fig.width=7, fig.height=7}
## Print the disparity_data object
disparity_data

## Summarising disparity in the different groups
summary(disparity_data)

## Plotting the results
plot(disparity_data)

## Testing for a difference between the groups
test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
```

# Specific tutorials

The following section contains information specific to some functions.
If any of your questions are not covered in these sections, please refer to the function help files in R, send me an email (<guillert@tcd.ie>), or raise an issue on GitHub.
The several tutorials below describe specific functionalities of certain functions; please always refer to the function help files for the full function documentation!

## Time slicing

The function `time.subsamples` allows users to divide the matrix into different time subsamples or slices given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.

Two types of time subsamples can be performed by using the `method` option:

 *  Discrete time subsamples (or time-binning) using `method = discrete`
 *  Continuous time subsamples (or time-slicing) using `method = continuous`

For the time-slicing method details see @GuillermeSTD. <!-- @@@ Change cite appropriately! -->
<!-- NC: Or potentially this paper we are writing for the PalAss? TG: totally!-->
For both methods, the function takes the `time` argument which can be a vector of `numeric` values for:

 *  Defining the boundaries of the time bins (when `method = discrete`)
 *  Defining the time slices (when `method = continuous`)

Otherwise, the `time` argument can be set as a single `numeric` value for automatically generating a given number of equidistant time-bins/slices.
Additionally, it is also possible to input a dataframe containing the first and last occurrence data (FAD/LAD) for taxa that span over a longer time than the given tips/nodes age, so taxa can appear in more than one time bin/slice.

Here is an example for `method = discrete`:

```{r, eval=TRUE}
## Generating three time bins containing the taxa present every 40 Ma
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = c(120, 80, 40, 0))
```

Note that we can also generate equivalent results by just telling the function that we want three time-bins as follow:

```{r, eval=TRUE}
## Automatically generate three equal length bins:
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = 3)
```

In this example, the taxa were split inside each time-bin according to their age.
However, the taxa here are considered as single points in time.
It is totally possible that some taxa could have had longer longevity and that they exist in multiple time bins.
In this case, it is possible to include them in more than one bin by providing a table of first and last occurrence dates (FAD/LAD).
This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:

```{r, eval=TRUE}
## Displaying the table of first and last occurrence dates for each taxa
head(BeckLee_ages)

## Generating time bins including taxa that might span between them
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
                time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
```

When using this method, the oldest boundary of the first bin (or the first slice, see below) is automatically generated as the root age plus 1\% of the tree length, as long as at least three elements/taxa are present at that point in time.
The algorithm adds an extra 1\% tree length until reaching the required minimum of three elements.
It is also possible to include nodes in each bin by using `inc.nodes = TRUE` and providing a matrix that contains the ordinated distance among tips *and* nodes.

For the time-slicing method (`method = continuous`), the idea is fairly similar.
This option, however, requires a matrix that contains the ordinated distance among taxa *and* nodes and an extra argument describing the assumed evolutionary model (via the `model` argument).
This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node, meaning that a decision must be made about what the value for the branch should be.
The model can be one of the following:

 *  `acctran` where the data chosen along the branch is always the one of the descendant
 *  `deltran` where the data chosen along the branch is always the one of the ancestor
 *  `punctuated` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
 *  `gradual` where the data chosen along the branch is either the descendant or the ancestor depending on branch length

```{r, eval=TRUE}
## Generating four time slices every 40 million years under a model of gradual evolution
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)

## Generating four time slices automatically
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
```

## Customised subsamples

Another way of separating elements into different categories is to use customised subsamples as briefly explained [above](#disparity-among-groups).
This function simply takes the list of elements to put in each group (whether they are the actual element names or their position in the matrix).

```{r, eval=TRUE}
## Creating the two groups as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44))

## Separating the dataset into two different groups
custom.subsamples(BeckLee_mat50, group = mammal_groups)
```

Elements can easily be assigned to different groups if necessary!

```{r, eval=FALSE}
## Creating the three groups as a list
mammal_groups <- list("crown" = c(16, 19:41, 45:50),
                      "stem" = c(1:15, 17:18, 42:44).
                      "all" = c(1:50))
```

## Bootstraps and rarefactions

One important step in analysing ordinated matrices is to pseudo-replicate the data to see how robust the results are, and how sensitive they are to outliers in the dataset.
This can be achieved using the function `boot.matrix` to bootstrap and/or rarefy the data.
The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):

```{r, eval=TRUE}
## Default bootstrapping
boot.matrix(data = BeckLee_mat50)
```

The number of bootstrap replicates can be defined using the `bootstraps` option.
The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
Currently two algorithms are implemented:

 * `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any *m* elements drawn from the data)
 * `single` where only one random element is replaced by one other random element for each pseudo-replicate

```{r, eval=TRUE}
## Bootstrapping with the single bootstrap method
boot.matrix(BeckLee_mat50, boot.type = "single")
```

This function also allows users to rarefy the data using the `rarefaction` argument.
Rarefaction allows users to limit the number of elements to be drawn at each bootstrap replication.
This is useful if, for example, one is interested in looking at the effect of reducing the number of elements on the results of an analysis.

This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsample to a minimum of three elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.

```{r, eval=TRUE}
## Bootstrapping with the full rarefaction
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)

## Or with a set number of rarefaction levels
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8, 3))
```

One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When missing, all dimensions from the ordinated matrix are used.

```{r, eval=TRUE}
## Using the first 50% of the dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

## Using the first 10 dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could directly supply the subsamples generated above (using `time.subsamples` or `custom.subsamples`) to this function.

```{r, eval=TRUE}
## Creating subsamples of crown and stem mammals
crown_stem <- custom.subsamples(BeckLee_mat50,
                                group = list("crown" = c(16, 19:41, 45:50), 
                                             "stem" = c(1:15, 17:18, 42:44)))
## Bootstrapping and rarefying these groups
boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)

## Creating time slice subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
                               method = "continuous", model = "gradual", 
                               time = c(120, 80, 40, 0),
                               FADLAD = BeckLee_ages)

## Bootstrapping the time slice subsamples
boot.matrix(time_slices, bootstraps = 100)
```

## Disparity metrics

There are many ways of measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space @DonohueDim, the sum and the product of the ranges and variances @Wills1994 or the median position of the elements relative to their centroid @Wills1994. 
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks, and some being more generalist.

Because of this great diversity of metrics, the package `dispRity` does not have one way to measure disparity but rather proposes to facilitate users in defining their own disparity metric that will best suit their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
However the `metric` argument has to follow certain rules:

1.  It must be composed from one to three `function` objects;
2.  The function(s) must take as a first argument a `matrix` or a `vector`;
3.  The function(s) must be of one of the three dimension-levels described below;
4.  At least one of the functions must be of dimension-level 1 or 2 (see below).

### The function dimension-levels

The metric function dimension-levels determine the "dimensionality of decomposition" of the input matrix.
In other words, each dimension-level designates the dimensions of the output, i.e. either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimension.

![Illustration of the different dimension-levels of functions with an input `matrix`](dispRity_fun.png)

#### Dimension-level 1 functions

A dimension-level 1 function will decompose a `matrix` or a `vector` into a single value:

```{r}
## Creating a dummy matrix
dummy_matrix <- matrix(rnorm(12), 4, 3)

## Example of dimension-level 1 functions
mean(dummy_matrix)
median(dummy_matrix)
```

Any summary metric such as mean or median are good examples of dimension-level 1 functions as they reduce the matrix to a single dimension (i.e. one value).

#### Dimension-level 2 functions

A dimension-level 2 function will decompose a `matrix` into a `vector`.

```{r}
## Defining the function as the product of rows
prod.rows <- function(matrix) apply(matrix, 1, prod)

## A dimension-level 2 metric
prod.rows(dummy_matrix)
```

Several dimension-level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` or `ranges` functions that calculate the variance or the range of each dimension of the ordinated matrix respectively.

#### Dimension-level 3 functions

Finally a dimension-level 3 function will transform the matrix into another matrix.
Note that the dimension of the output matrix doesn't need to match the the input matrix:

```{r}
## A dimension-level 3 metric
var(dummy_matrix)

## A dimension-level 3 metric with a forced matrix output
as.matrix(dist(dummy_matrix))
```

### `make.metric`

Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the Euclidean distance between each element and the centroid of the ordinated space.
The `make.metric` function implemented in `dispRity` is designed to help test and find the dimension-level of the functions.
This function tests:

1.  If your function can deal with a `matrix` or a `vector` as an input;
2.  Your function's dimension-level according to its output (dimension-level 1, 2 or 3, see above);
3.  Whether the function can be implemented in the `dispRity` function (the function is fed into a `lapply` loop).

For example, let's see if the functions described above are the right dimension-levels:

```{r}
## Which dimension-level is the mean function? And can it be used in dispRity?
make.metric(mean)

## Which dimension-level is the prod.rows function? And can it be used in dispRity?
make.metric(prod.rows)

## Which dimension-level is the var function? And can it be used in dispRity?
make.metric(var)
```

A non verbose version of the function is also available.
This can be done using the option `silent = TRUE` and will simply output the dimension-level of the metric.

```{r}
## Testing whether mean is dimension-level 1
if(make.metric(mean, silent = TRUE) != "level1") {
    message("The metric is not dimension-level 1.")
}
## Testing whether var is dimension-level 1
if(make.metric(var, silent = TRUE) != "level1") {
    message("The metric is not dimension-level 1.")
}
```

### Metrics in the `dispRity` function

Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:

```{r}
## Measuring disparity as the standard deviation of all the values of the
## ordinated matrix (dimension-level 1 function).
summary(dispRity(BeckLee_mat50, metric = sd))

## Measuring disparity as the standard deviation of the variance of each axis of
## the ordinated matrix (dimension-level 1 and 2 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))

## Measuring disparity as the standard deviation of the variance of each axis of
## the variance covariance matrix (dimension-level 1, 2 and 3 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function dimension-levels (using `make.metric`) and apply them to the data in decreasing order (dimension-level 3 > 2 > 1).

```{r}
## Disparity as the standard deviation of the variance of each axis of the
## variance covariance matrix:
disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)),
                      round = 10)

## Same as above but using a different function order for the metric argument
disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)),
                      round = 10)

## Both ways output the same disparity values:
disparity1 == disparity2
```

In these examples, we considered disparity to be a single value.
For example, in the previous example, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
It is, however, possible to calculate [disparity as a distribution](#disparity-as-a-distribution).

### Metrics implemented in `dispRity`

Several disparity metrics are implemented in the `dispRity` package.
The detailed list can be found in `?dispRity.metric` along with some description of each metric.

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `variances` | The variance of each dimension | `dispRity` |
2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `mode.val` | The modal value | `dispRity` |

1: This function uses an estimation of the eigenvalue that only works for MDS or PCoA ordinations (not PCA).

2: Note that by default, the centroid is the centroid of the elements.
It can, however, be fixed to a different value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.

### Equations and implementations
Some of the functions described below are implemented in the `dispRity` package and do not require any other packages to calculate ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}
Where *k* is the number of dimensions, and $\lambda_i$ is the eigenvalue of each dimension.

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}
Where *k* is the number of dimensions.

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}
Where *k* is the number of dimensions.

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}
Where *k* is the number of dimensions, and $\sigma^{2}$ is their variance.

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}
Where *n* is each element in the ordinated space, *k* is the number of dimensions, and $Centroid_{k}$ is their mean (or can be set to another value).

### Using the different disparity metrics
Here is a brief demonstration of the main metrics implemented in `dispRity`.
First, we will create a dummy/simulated ordinated space using the `space.maker` utility function (more about that [here](#space.maker):

```{r}
## Creating a 10*5 normal space
set.seed(1)
dummy_space <- space.maker(10, 5, rnorm)
```

We will use this simulated space to demonstrate the different metrics.

#### Volumes and surface metrics
The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupied:

```{r}
## Calculating the ellipsoid volume
summary(dispRity(dummy_space, metric = ellipse.volume))
```
> Because there is only one subsample (i.e. one matrix) in the dispRity object, this operation is the equivalent of `ellipse.volume(dummy_space)` (with rounding).

```{r}
## Calculating the convex hull surface
summary(dispRity(dummy_space, metric = convhull.surface))

## Calculating the convex hull volume
summary(dispRity(dummy_space, metric = convhull.volume))
```

The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
For the full and correct calculation of the volume of the ordinated space, it is possible to use the `hyper.volume` function that has more options but takes longer to calculate.

```{r}
## Calculating the true multidimensional volume
summary(dispRity(dummy_space, metric = hyper.volume))

```
> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) that often results in near 0 disparity values.

#### Ranges, variances and diagonal

The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space based on its dimensional properties (they are also less affected by the "curse of dimensionality"):

`ranges` and `variances` both work on the same principle and measure the range/variance of each dimension:

```{r}
## Calculating the ranges of each dimension in the ordinated space
ranges(dummy_space)

## Calculating disparity as the distribution of these ranges
summary(dispRity(dummy_space, metric = ranges))

## Calculating disparity as the sum and the product of these ranges
summary(dispRity(dummy_space, metric = c(sum, ranges)))
summary(dispRity(dummy_space, metric = c(prod, ranges)))

## Calculating the variances of each dimension in the ordinated space
variances(dummy_space)

## Calculating disparity as the distribution of these variances
summary(dispRity(dummy_space, metric = variances))

## Calculating disparity as the sum and the product of these variances
summary(dispRity(dummy_space, metric = c(sum, variances)))
summary(dispRity(dummy_space, metric = c(prod, variances)))
```

The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest Euclidean distance in our five dimensional space):

```{r}
## Calculating the ordinated space's diagonal
summary(dispRity(dummy_space, metric = diagonal))
```

> This metric is only a Euclidean diagonal (mathematically valid) if the dimensions within the space are all orthogonal!

#### Centroids metric

The `centroids` metric allows users to measure the position of the different elements compared to a fixed point in the ordinated space.
By default, this function measures the distance between each element and their centroid (centre point):

```{r}
## The distribution of the distances between each element and their centroid
summary(dispRity(dummy_space, metric = centroids))

## Disparity as the median value of these distances
summary(dispRity(dummy_space, metric = c(median, centroids)))
```

It is however possible to fix the coordinates of the centroid to a specific point in the ordinated space, as long as it has the correct number of dimensions:

```{r}
## The distance between each element and the origin of the ordinated space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))

## Disparity as the distance between each element and a specific point in space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
```

## Summarising dispRity data (plots)

Because of its architecture, printing `dispRity` objects only summarises their content but does not print the disparity value measured or associated analysis (more about this [here](#manipulating-dispRity-objects)).
To actually see what is in a dispRity object, one can either use the `summary` function for visualising the data in a table or `plot` to have a graphical representation of the results.

### Summarising `dispRity` data
This function is an S3 function (`summary.dispRity`) allowing users to summarise the content of `dispRity` objects that contain disparity calculations.

<!-- DATA LOADER -->

```{r}
## Example data from previous sections
crown_stem <- custom.subsamples(BeckLee_mat50,
                                group = list("crown" = c(16, 19:41, 45:50), 
                                             "stem" = c(1:15, 17:18, 42:44)))
## Bootstrapping and rarefying these groups
boot_crown_stem <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE)
## Calculate disparity
disparity_crown_stem <- dispRity(boot_crown_stem, metric = c(sum, variances))

## Creating time slice subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping the time slice subsamples
boot_time_slices <- boot.matrix(time_slices, bootstraps = 100)
## Calculate disparity
disparity_time_slices <- dispRity(boot_time_slices, metric = c(sum, variances))

## Creating time bin subsamples
time_bins <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages,
    inc.nodes = TRUE)
## Bootstrapping the time bin subsamples
boot_time_bins <- boot.matrix(time_bins, bootstraps = 100)
## Calculate disparity
disparity_time_bins <- dispRity(boot_time_bins, metric = c(sum, variances))
```

These objects are easy to summarise as follows:

```{r}
## Default summary
summary(disparity_time_slices)
```

Information about the number of elements in each subsample and the observed (i.e. non-bootstrapped) disparity are also calculated.
This is specifically handy when rarefying the data for example:

```{r}
head(summary(disparity_crown_stem))
```

The summary functions can also take various options such as:

 * `quantile` values for the confidence interval levels (by default, the 50 and 95 quantiles are calculated)
 * `cent.tend` for the central tendency to use for summarising the results (default is `median`)
 * rounding` option corresponding to the number of decimal places to print (default is `2`)
 * `recall` option for printing the call of the `dispRity` object as well (default is `FALSE`)

These options can easily be changed from the defaults as follows:

```{r}
## Same as above but using the 88th quantile and the standard deviation as the summary 
summary(disparity_time_slices, quantile = 88, cent.tend = sd)

## Printing the details of the object and rounding the values to the 5th decimal place
summary(disparity_time_slices, recall = TRUE, rounding = 5)
```

Note that the summary table is a `data.frame`, hence it is as easy to modify as any dataframe using `dplyr`.
You can also export it in `csv` format using `write.csv` or `write_csv` or even directly export into `LaTeX` format using the following;

```{r, eval=FALSE}
## Loading the xtable package
require(xtable)
## Converting the table in LaTeX
xtable(summary(disparity_time_slices))
```

### Plotting `dispRity` data
An alternative (and more fun!) way to display the calculated disparity is to plot the results using the S3 method `plot.dispRity`.
This function takes the same options as `summary.dispRity` along with various graphical options described in the function help files (see `?plot.dispRity`).

The plots can be of four different types:

 * `continuous` for displaying continuous disparity curves
 * `box`, `lines`, and `polygons` to display discrete disparity results in respectively a boxplot, confidence interval lines, and confidence interval polygons.

> This argument can be left empty. In this case, the algorithm will automatically detect the type of subsamples from the `dispRity` object and plot accordingly.

It is also possible to display the number of elements in each subsample (as a horizontal dotted line) using the option `elements = TRUE`.
Additionally, when the data is rarefied, one can indicate which level of rarefaction to display (i.e. only display the results for a certain number of elements) by using the `rarefaction` argument.

```{r, fig.width=8, fig.height=8}
## Graphical parameters
op <- par(mfrow = c(2, 2), bty = "n")

## Plotting continuous disparity results
plot(disparity_time_slices, type = "continuous")

## Plotting discrete disparity results
plot(disparity_crown_stem, type = "box")

## As above but using lines for the rarefaction level of 20 elements only
plot(disparity_crown_stem, type = "line", rarefaction = 20)

## As above but using polygons while also displaying the number of elements
plot(disparity_crown_stem, type = "polygon", elements = TRUE)

## Resetting graphical parameters
par(op)
```

Since `plot.dispRity` uses the arguments from the generic `plot` method, it is of course possible to change pretty much everything using the regular plot arguments:

```{r, fig.width=8, fig.height=8}
## Graphical options
op <- par(bty = "n")

## Plotting the results with some classic options from plot
plot(disparity_time_slices, col = c("blue", "orange", "green"),
    ylab = c("Some measurement"), xlab = "Some other measurement",
    main = "Many options...", ylim = c(5, 0), xlim = c(4, 0))

## Adding a legend
legend("topleft", legend = c("Central tendency",
                             "Confidence interval 1",
                            "Confidence interval 2"),
      col = c("blue", "orange", "green"), pch = 19)

## Resetting graphical parameters
par(op)
```

In addition to the classic `plot` arguments, the function can also take arguments that are specific to `plot.dispRity` like adding the number of elements or rarefaction level (as described above), and also changing the values of the quantiles to plot as well as the central tendency.

```{r, fig.width=8, fig.height=8}
## Graphical options
op <- par(bty = "n")

## Plotting the results with some plot.dispRity arguments
plot(disparity_time_slices, quantile = c(seq(from = 10, to = 100, by = 10)),
    cent.tend = sd, type = "c", elements = TRUE, col = c("black", rainbow(10)),
    ylab = c("Disparity", "Diversity"), time.subsamples = FALSE,
    xlab = "Time (in in units from past to present)", observed = TRUE,
    main = "Many more options...")

## Resetting graphical parameters
par(op)
```

> Note that the argument `observed = TRUE` allows to plot the disparity values calculated from the non-bootstrapped data as crosses on the plot.

For comparing results, it is also possible to add a plot to the existent plot by using `add = TRUE`:

```{r, fig.width=8, fig.height=8}
## Graphical options
op <- par(bty = "n")

## Plotting the continuous disparity with a fixed y axis
plot(disparity_time_slices, ylim = c(1, 4))
## Adding the discrete data
plot(disparity_time_bins, type = "line", ylim = c(1, 4), xlab = "", ylab = "",
    add = TRUE)

## Resetting graphical parameters
par(op)
```

Finally, if your data has been fully rarefied, it is also possible to easily look at rarefaction curves by using the `rarefaction = TRUE` argument:

```{r, fig.width=8, fig.height=8}
## Graphical options
op <- par(bty = "n")

>>>>

--- 
title: "dispRity manual"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: TGuillerme/dispRity
description: "dispRity R package vignette"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

@@
